---
title: "DEMO : Spinal Cord Axis Projection Example"
output: 
  tufte::tufte_html: default
author: Alma Andersson
date: 2021-10-01
---

## Prep

This notebook illustrates how the distribution of certain features of interest
(e.g., genes or cell types) along a specified axis can be analyzed. To simplify
things, I've written the code as a small package that you can load functions
from. These functions are written to be versatile and general, as to allow
custom analysis of different kinds.

**NOTE** : This is just a demo and not meant to be a comprehensive discussion on
the actual code.

The first thing you need to do is to specify the paths to your data
(`DATA_PATH`):


```{r}
DATA_PATH <- "../../data/robjs/210902_SC_DVaxis.rds"
```

Unsurprisingly, the next thing we do is to load the necessary packages (mainly
so that we can read your `Seurat` object) and source the custom functions.

```{r}
library(Seurat)
library(ggplot2)
library(gridExtra)
```

We proceed to install and load the package I've written for you. We fetch this
package from `github`:

```{r}
if (!("axis.projection" %in% rownames(installed.packages()))){
   devtools::install_github("git@github.com:almaan/axis-projection.git") 
}

library(axis.projection)
```

## Data loading and processing

We will now load the data that we want to analyze (your `Seurat` object), and
make some small adjustments to make it easier to analyze it.

```{r}
se <- readRDS(DATA_PATH)
head(se@meta.data[c("labels","axis_annot")])
```
While it's nice that the column `axis_annot` in the metadata also contains the
section information, this complicates things a bit when we want to identify the
DV axis, the main issue being that a spit annotated as `1_D` would be interpret
as something different than for example `2_D`, when both spots should be
considered as _dorsal_. Also, there were some inconsistencies in the naming of
these sections (different arrangement of "_" and numbers).  Hence, we'll adjust
this before proceeding - the updated annotations will be placed in a new column
called `DV`, where we only have the "D" and "V" annotation.

```{r}
# create a DV column
se@meta.data["DV"] <- gsub("[1-9]|_","",se@meta.data$axis_annot)

```

Since this is only a demo, we'll only use 4 of the sections you've identified in
the `labels` column, but of course the same workflow can be applied to the complete 
data set. I've written the code to be memory efficient, so it should not be any
issues increasing the amount of data.

```{r}
# list which sections you want to keep
keep.sections <- paste0("section_",c(12,15,19,23))
# print info
print(sprintf("Well keep the sections >> %s",paste(keep.sections,collapse=", ")))
# select spots belonging to sections
keep.spots <- which(se@meta.data$labels %in% keep.sections)
# subset data, make sure to use the SubsetSTData to correctly modify Staffli object
se <- STutility::SubsetSTData(se,spots = keep.spots)
# inspect new object
se
```

## Analysis : Axis projection

We may now actually analyze the data. In order to do this we must do
two things:

1. From knowing which spots that are in the _dorsal_ respectively _ventral_
parts, we need to computationally **identify the DV-axis**. The way I've
approached this issue to, for every spot marked as _ventral_ compute the vector
that connects it with it's nearest _dorsal_ spot. Once these vectors are
computed, they are normalized to unit norm (euclidean) and the mean is computed.
From this we should end up with a fairly good estimate of the axis vector.

2. Once the DV-axis is identified, we want to project every spot onto the
new space spanned by the axis vector, and extract the coordinates w.r.t. this 1D
space. For this I've used the standard projection formula:

$proj_{\bar{v}}\bar{x} = \frac{\bar{x}\cdot\bar{v}}{\bar{v}\cdot\bar{v}}\bar{v}$

Here $\cdot$ represents the dot-product and $\bar{x}$ is the coordinates of a
spot to be projected on the DV-axis ($\bar{v}$). Normalizing $\bar{v}$ to unit
norm, simplifies the formula to:

$proj_{\bar{v}}\bar{x} = (\bar{x}\cdot\bar{v})\bar{v}$

from which it's evident that the projection's coordinate w.r.t. the basis $\bar{v}$ is simply:
$\bar{x}\cdot\bar{v}$

Both of these steps, will be automatically performed by the function
`axis.projection` that I wrote and included in the package. This function takes
3 required and 1 optional argument. In context of this function we'll be
speaking of "A" and "B" annotations/labels, where the idea is that the axis goes
from "A->B". These arguments are:

- `ab.column` : column in the object's `meta.data` slot that holds the
   annotations of whether the spots belongs to A or B. _In your case this is_ `DV`
- `a.label` : label used to indicate that a spot belongs to A. _In your case this is_ `V`
- `b.label` : same as `a.label` but for B. In your case this is `D`.
- `split.on` : if an individual axis should be computed for subset of spots,
   then `split.on` should be set to the column name that assigns the spots to
   each of these subsets. _In your case_ this should be `labels`, since you need an individual axis for each section.
-  `include.orthogonal` : By setting this to true, we also get the projections on the axis orthogonal to the ab-axis.

```{r}
# set argument values
ab.column <- "DV"
a.label <- "V"
b.label <- "D"
split.on <- "labels"
# indentify axis and project data on it
se <- axis.projection(se,
                      ab.column = ab.column,
                      a.label = a.label,
                      b.label = b.label,
                      split.on = split.on,
                      include.orthogonal = T
                      )

```

As another sanity check, we can plot the sections and color the spots according to
their position along the axis (**note**, so far this has nothing to do with
expression values). Here, a low value means that the spot is near the ventral
edge while a high value means the spot is near the dorsal edge.

```{r,fig.width=10}

# create empty list for grobs
grobs <- list()
# iterate over each section
for (s in keep.sections){
  
  # select spot for section
  test.spots <- which(se@meta.data$labels == s)
  # select coordinates for section
  crd <- get.coordinates(STutility::SubsetSTData(se,spots = test.spots))
  # get projected data
  proj.val <- se@meta.data$DV_projection[test.spots]
#  proj.val <- se@meta.data$orth_to_DV_projection[test.spots]
  
  # create data frame for plotting
  test.plot.df <- data.frame(x= crd$pixel_x,
                             y = crd$pixel_y,
                             projval = proj.val
                             ) 
  
  # create plot and store in list
  grobs[[s]] <- ggplot(test.plot.df, aes(x=x,
                                         y = y,
                                         color = projval,
                                         fill = projval)) +
                geom_point(size = 2) +
                coord_fixed() +
                scale_y_reverse()
}  

# visualize subplots in grid
grid.arrange(grobs = grobs,
             ncol = 2)

```

That looks good, seems like the linear algebra implementations in the
`axis.projection` function works as expected. 

This means that we're now ready to actually look at the distribution of a
feature of interest along this axis.

### Visualization

The functions in the package `axis projection` are written to extract
information from the _default_ assay in you _Seurat_ object, so make sure to
that you have the correct value set here. For example, I want the `SCT`
normalized data, and hence we do:

```{r}
# set default assay to SCT
Seurat::DefaultAssay(object = se) <- "SCT"
```

Next, let us specify the features we want to investigate:

```{r}
# specify genes to investigate
features <- c("EBF1", "ZIC2","FGFBP3")
```

To visualize the collective distribution along the DV-axis across all four
samples we do:

```{r}
# visualize features along axis
g <- plot.axis.projection(se,
                          features,
                          ab.column,
                          a.label,
                          b.label,
                          normalize = T
                          )
print(g)
```
As we can see here, the expression levels of _EBF1_ and _ZIC2_ increase the
closer to the dorsal edge that we get, while the converse is true for _FGFBP3_.

We could also split the curves based on section, which might be handy when
looking at how these expression patterns change over time. To do this, all you
need to do is to _provide the name of the column you want to split the data with respect to_  (for you `labels`).

For example:

```{r}
# visualize features along axis
g <- plot.axis.projection(se,
                          features,
                          ab.column,
                          a.label,
                          b.label,
                          normalize = T,
                          split.on = "labels"
                          )
print(g)

```

Here features will have the same line color while the attribute you split w.r.t
have a different fill color. With 3 genes and 4 samples we get 12 curves in
total, which might be a bit messy. To circumvent this mess you can simply split
the results into several subplots. Also, the observed values (to which we have
fitted the smoothed curves), can be excluded from the visualization by setting
`show.data = FALSE`.

```{r}

# visualize features along axis
# one subplot for each feature
grobs <- lapply(features,
                function(feature){plot.axis.projection(se,
                          feature,
                          ab.column,
                          a.label,
                          b.label,
                          show.data =F,
                          normalize = T,
                          title = feature,
                          split.on = "labels"
                          )})

# visualize in grid
grid.arrange(grobs = grobs,
             ncol = 2)


```
You could also use different color palettes for the color and fill, to do this
just provide these as values to the arguments `color.color` and `fill.color`. If
you want to change the x-and ylabels, you can use the `xlabel` and `ylabel`
arguments.

```{r}
# visualize 
g <- plot.axis.projection(se,
                          features[1],
                          ab.column,
                          a.label,
                          b.label,
                          normalize = T,
                          split.on = "labels",
                          color.color = scale_color_hue(),
                          fill.color = scale_color_hue(),
                          xlabel = "Axis",
                          ylabel = "Gene Expression"
                          )

print(g)

```


## Finding spatially informative genes

Next thing for us to do is to try and extract a set of genes that are highly
informative of the spatial position along the DV-axis and the axis orthogonal to
this one.

The way we'll do this is by penalized logistic regression (more specifically we
apply a form of L2 regression), where we try to predict the position along each
axis from the gene expression. Although logistic regression is often applied to
Bernoulli distributed variables, it can also be applied to continous variables
like our axis positions. Logistic regression is also fit for this problem since
the logit function maps values on the real number line to the interval $(0,1)$,
which is exactly the values our axis-projections takes.

The generalized linear model (GLM) we are using for the fitting doesn't scale
that well to large datasets with plenty of features (like ours), so I've
implemented a form of binning where samples that has approximately the same
axis-coordinates are joined together and average expression profiles for them
are calculated. By binning the data, we can make use of all the information but
still keep the time for the analysis within reasonable time.

All the machinery that is needed to extract a set of spatially informative genes (SIGs)
is spread across several functions that have been strung together in the
`get.axis.genes` function. You can change the number of bins as well as the
number of genes to extract by the arguments `n.bins` and `n.genes` respectively.


Here we'll extract $1500$ genes associated with the DV-axis and $1500$ genes
associated with the axis that is orthogonal to the DV-axis, and then join them
together.


```{r}
set.seed(1337)

y.var.dv <- se@meta.data$DV_projection
y.var.dvo <- se@meta.data$orth_to_DV_projection

X.var <- t(as.matrix(se@assays$SCT[,]))

dv.axis.genes <- get.axis.genes(X.var,
                                y.var.dv,
                                bin.data = T,
                                n.bins = 100,
                                n.genes = 1500)
dvo.axis.genes <- get.axis.genes(X.var,
                                 y.var.dvo,
                                 bin.data = T,
                                 n.bins = 100,
                                 n.genes = 1500)

axis.genes <- unique(c(dv.axis.genes,dvo.axis.genes))
var.genes <- se@assays$SCT@var.features
print(sprintf("SI and HVG gene info | [HVGs] : %d | [SIGs] : %d | [Intersection] : %d ",length(var.genes),
              length(axis.genes),length(intersect(var.genes,axis.genes))))

```
As you can see there's quite a difference between the two gene sets, I would recommend you to try to:
1. Conduct the analysis with only the SIG set
2. Conduct the analysis with the union of the HVG and SIG sets


Also, as a sanity check we can see if the SIG genes actually seem to exhibit a
spatial character. Let us plot the top 5 (highest ranked) SIG genes associated
with the DV-axis.

```{r}
# visualize features along axis
g <- plot.axis.projection(se,
                          dv.axis.genes[1:5],
                          ab.column,
                          a.label,
                          b.label,
                          normalize = T
                          )
print(g)
```
Seems like we found some nice patterns there!


## Final Remarks

This is just a brief demo on how to use the functions I've written for you, let
me know if there are any questions. If you're interested in the actual code
itself, have a look at
[this](https://github.com/almaan/axis-projection/tree/master/R).

